---
title: 'Neural Networks: Heart Disease'
author: "Data Mining"
output: pdf_document
---

## Data

In this exercise we are going to analyze coronary heart disease determinants. We use data taken from a larger
dataset, described in  Rousseauw et al, 1983, South African Medical Journal. It contains measurements for several variables of males in a heart-disease high-risk region of the Western Cape, South Africa. These data are, amomg others,

Variable    | Description 
------------| -------------
*sbp*   		|systolic blood pressure
*tobacco*		|cumulative tobacco (kg)
*ldl*	    	|low density lipoprotein cholesterol
*famhist*		|family history of heart disease (Present, Absent)
*alcohol*		|current alcohol consumption
*age*		    |age at onset
*chd*		    |response, coronary heart disease

The data is available online and can be loaded directly into R.

```{r}
heart = read.table("https://web.stanford.edu/~hastie/ElemStatLearn/datasets/SAheart.data",
                 sep=",",head=T,row.names=1)
attach(heart)
```

We scale the explanatory variables by defining

$$x_{scale} = \frac{x-\min{x}}{\max{x}-\min{x}}.$$

Furthermore, we create a data frame with the scaled explanatory variables and the dependent variable.

```{r}
raw = data.frame(tobacco,adiposity,age,ldl)
max = apply(raw , 2 , max)
min = apply(raw , 2 , min)
scaled = as.data.frame(scale(raw, center = min, scale = max - min))

data = data.frame(chd,scaled)
```

## Logistic Regression

As benchmark, we estimate the model using logistic regression.


```{r}
glm.fit = glm(chd~tobacco+ldl+age+adiposity,family = binomial)
```

And predict the classes.

```{r}
glm.pred = predict(glm.fit,type="response")
chd.pred.lr = 1*(glm.pred>0.5)
```

We obtain the confusion matrix.

```{r}
tab.lr = table(chd,chd.pred.lr)
tab.lr
```

And we obtain the false positive and true positive rates.

```{r}
fpr.lr = tab.lr[1,2]/length(chd[chd==0])
tpr.lr = tab.lr[2,2]/length(chd[chd==1])

fpr.lr
tpr.lr
```

## Neural Networks

We estimate a neural network using the *neuralnet* function from the package of the same name.

We decide to use a network with two hidden layers with three and two neurons, respectively.

```{r}
library(neuralnet)
nn.heart = neuralnet(chd~tobacco+ldl+age+adiposity,data=data,hidden = c(3,2))
```

We plot the neural network just estimated.

```{r}
plot(nn.heart)
```

We use the estimated neural network top predict the classes. We use a 0.5 threshold.

```{r}
predict.nn = compute(nn.heart, data)
chd.pred.nn = 1*(predict.nn$net.result>0.5)
```

Obtaining the confusion matrix.

```{r}
tab.nn = table(chd,chd.pred.nn)
tab.nn
```

And computing the false positive and true positive rates.

```{r}
fpr.nn = tab.nn[1,2]/length(chd[chd==0])
tpr.nn = tab.nn[2,2]/length(chd[chd==1])

fpr.nn
tpr.nn
```
